A White Noise Approach to Evolutionary Ecology

### 2020-04-28 Current Version/Untitled.tex
reorganize

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={A White Noise Approach to Evolutionary Ecology},
            pdfauthor={Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{A White Noise Approach to Evolutionary Ecology}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{3/1/2020}

\usepackage{float}
\usepackage{flafter}

\usepackage{xr}
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{hyperref}
\usepackage[left]{lineno}
\usepackage{amsmath}
\linenumbers
\externaldocument{SM}

\begin{document}
\maketitle
\begin{abstract}
We derive the dynamics of the distribution of a quantitative character
and the abundance of a biological population from a stochastic partial
differential equation driven by space-time white noise. In the process
we develop a useful set of heuristics to operationalize the powerful,
but abstract theory of white noise and measure-valued stochastic
processes. This approach allows us to compute the full implications of
demographic stochasticity on phenotypic distributions and abundances of
populations. We demonstrate the utility of our approach by deriving a
quantitative genetic model of diffuse coevolution mediated by
exploitative competition for a continuum of resources. Other than trait
and abundance distributions, this model predicts interaction networks
parameterized by rates of interactions, competition coefficients, and
selection gradients. We briefly investigate the relationship between
selection gradients and competition coefficients. This illustrative
investigation suggests absolute values of linear selection gradients are
negatively correlated with competition coefficients when the ratio of
interspecific trait variation to intraspecific trait variation is
approximately less than \(1.25\) and positively correlated otherwise.
The inverse relationship holds for quadratic selection gradients except
with threshold diversity ratio equal to unity instead of \(1.25\).
Hence, this approach contributes to the development of a synthetic
theory of evolutionary ecology by formalizing first principle
derivations of stochastic models that underlie rigorous investigations
of the relationship between feedbacks of biological processes and the
patterns of diversity they produce.
\end{abstract}

@@

Our goal in this manuscript is to develop a rigorous, but accessible
approach to synthesize the stochastic dynamics of abundance, mean trait
and heritable variation in biological populations for the study of
theoretical evolutionary ecology. A primary aim of theoretical
evolutionary ecology is the development of mathematical approaches to
describe the evolution of populations and their interactions with both
the biotic and abiotic environments in which they are embedded. Given
this consideration, a natural scope for such an approach centers on
quantifying the abundance dynamics of populations and the evolution of
traits mediating their interactions as functions of relevant abiotic
factors. Although taking into account abundance, phenotype and
environment provides the basis for a partial understanding of the
complex nature of biological communities, a deeper understanding must
account for the effects of dispersal and the phylogeographic history of
interacting lineages (Kraft et al. 2007; Hickerson et al. 2010; Manceau,
Lambert, and Morlon 2016; McPeek 2017) along with the genetic basis of
ecologically relevant traits (Conner 2004; Fussman, Loreau, and Abrams
2007) and feedbacks between populations and the biogeochemical cycles
they ultimately depend on (Loreau 2010; Ågren and Andersson 2012). It is
therefore ideal that the development of any such mathematical approach
anticipates extensions to account for these important factors shaping
ecological communities, especially as empirical and conceptual work in
these directions continues to grow (Abdala-Roberts and Mooney 2014;
Kölzsch et al. 2015; Crutsinger 2015; Fitzpatrick et al. 2015, 2017;
Marx et al. 2017; Rudman et al. 2017; Skovmand et al. 2018; Nuland et
al. 2019; Harmon et al. 2019). Furthermore, the approach would benefit
from a stochastic component to capture the chance nature of biological
reality (Lande, Engen, and SÆther 2003; Meester et al. 2018; Mubayi et
al. 2019) and serve as a basis for the construction of statistical
methods that measure evolutionary and ecological processes occuring in
the wild. Such methods will tether theory to reality and allow for
rigorous tests of hypotheses on the structure and behavior of ecological
communities. In this paper we introduce a framework that establishes a
formal connection between the continuous-time dynamics of abundance and
quantitative traits in stochastically evolving populations. We then
demonstrate the utility of our framework through the derivation and
analysis of a model of diffuse coevolution and discuss how it can be
extended to account for the details mentioned above.

Current theoretical approaches to synthesize evolution and ecology have
capitalized on the fact that biological fitness plays a key role in
determining both sets of dynamics. While correlation of fitness and
genotype is the basis of evolution by natural selection, the mean
fitness across all individuals in a population determines the growth,
stasis or decline of abundance. In section \ref{deterministic} we review
the mathematical formalization of this connection, which has been
established in the contexts of population genetics (Crow and Kimura
1970; Roughgarden 1979), evolutionary game theory (Hofbauer and Sigmund
1998; Nowak 2006; Lion 2018), quantitative genetics (Lande 1982; Doebeli
1996; Lion 2018) and a unifying framework for these three distinct
approaches to evolutionary theory (Champagnat, Ferrière, and Méléard
2006) which includes our approach as a special case. We note here there
are two primary differences between the results presented in this paper
and those of Champagnat, Ferrière and Méléard (2006). First, instead of
developing a unifying approach to evolutionary theory, we focus on
developing a stochastic synthesis of population dynamics and
quantitative genetics including specific expressions for the dynamics of
abundance, mean trait and additive genetic variance. Second, although
Champagnat, Ferrière and Méléard (2006) provide rigorous proofs of
results relevant to our approach, they do not communicate the nessecary
tools we make use of for deriving the dynamics of abundance, mean trait
and additive genetic variance. We aim to communicate these tools to a
wide audience by striking a balance between intuitive appeal and
mathematical rigor.

Although reviewing previously established theoretical results reveals a
beautiful synthesis of evolution and population ecology, it also reveals
a gap in theoretical approaches to incorporate the intrinsically random
nature of populations. Specifically, in theoretical quantitative
genetics the derivation of a population's response to random genetic
drift is derived in discrete time under the assumption of constant
effective population size using arguments based on properties of random
samples (Lande 1976). Though this approach conveniently mimics the
formalism provided by the Wright-Fisher model of population genetics,
real population sizes fluctuate over time. Furthermore, since these
fluctuations are themselves stochastic, it seems natural to derive
expressions for the evolutionary response to demographic stochasticity
and consider how the results relate to characterizations of random
genetic drift. This can be done in continuous time for population
genetic models without too much technical overhead, assuming a finite
number of alleles (Lande, Engen, and Sæther 2009; Parsons, Quince, and
Plotkin 2010; Gomulkiewicz, Krone, and Remien 2017). However, for
populations with a continuum of types, such as a quantitative trait,
finding a formal approach to derive the evolutionary response to
demographic stochasticity has remained a vexing mathematical challenge.
In this paper we close this gap by combining the calculus of white noise
with results on rescaled limits of branching Brownian motion processes
(BBM) and stochastic partial differential equations (SPDE), which are
stochastic analogs of deterministic partial differential equations
(PDE). Our goal has two components: 1) Establish a novel synthetic
approach to theoretical evolutionary ecology that provides a formal
connection between demographic stochasticity and random genetic drift in
the context of quantitative traits. 2) Communicate some useful
properties of space-time white noise, BBM and SPDE to as wide of
audience as possible. With this goal in mind we will not provide a
rigorous treatment of any of these deep subjects. Instead, we introduce
a set of heuristics that only require the basic concepts of Reimann
integration, partial differentiation and some exposure to Brownian
motion and stochastic ordinary differential equations (SDE). For a
concise introduction to SDE and Brownian motion, we recommend the primer
by Evans (2014). Rigorous treatments of SPDE and rescaled limits of BBM
can be found in Walsh (1986) and Dawson (1993) respectively.

Through accomplishing these goals this work yields five main results.
Firstly, we develop a concise, but accurate approach for understanding
SPDE and provide heuristics to empower mathematical evolutionary
ecologists to derive novel models that relax key assumptions commonly
made in combining models of evolutionary and abundance dynamics. Second,
we formally derive general SDE driven by demographic stochasticity that
express the joint dynamics of abundance, mean trait and additive genetic
variance for populations exhibiting a wide range of potential phenotypic
distributions. Although these dynamics have been considered before while
accounting for demographic stochasticity (Polechová and Barton 2015), to
our knowledge this is the first time they have been rigorously derived
from first principles. Third, we simplify these SDE by approximating the
abundance density of phenotypes with a Gaussian curve and integrate the
results with a model of imperfect inheritance to find formula that
generalize classical quantitative genetic expressions. This includes an
expression for the response of additive genetic variance to selection in
terms of fitness gradients analogous to the expression for the response
in mean phenotype found by Lande (1976). Fourth, following classical
niche theory we provide a rigorous mechanistic derivation of fitness for
populations experiencing resource competition and abiotic stabilizing
selection using diffusion limits of measure-valued branching processes.
This calculation provides a demonstration for performing such diffusion
limits that can be applied in other contexts. Fifth, we explore
implications of the resulting model of diffuse coevolution by
investigating the relationship between selection gradients and
competition coefficients, discovering the conditions under which these
quantities positively or negatively covary.

@@

At the core of our approach is a stochastic analog of the replicator
equation with mutation in continuous time and phenotypic space (Taylor
and Jonker 1978; Schuster and Sigmund 1983). From this stochastic
replicator-mutator equation we derive a system of SDE for the dynamics
of abundance, mean trait and additive genetic variance of a population.
Hence, our approach develops a quantitative genetic theory of
evolutionary ecology. A popular alternative to quantitative genetics is
the theory of adaptive dynamics. As demonstrated by Page and Nowak
(2002) and Champagnat, Ferrière and Méléard (2006), the canonical
equation of adaptive dynamics can be derived from the replicator-mutator
equation. Thus, one could start from the atomic roots of our approach
and pursue a stochastic adaptive dynamic theory instead. We choose the
former in anticipation of an extension of our approach that explicitly
models the genetic details of populations.

@@

Our review begins by considering the dynamics of an asexually
reproducing population in a homogeneous environment. For simplicity, we
first assume individuals are haploid and carry one of \(K\) alleles each
with a different fitness expressed as growth rate before introducing a
model involving a quantitative trait. Under these assumptions, the
derivation of the evolution of allele frequencies due to natural
selection can be derived from expressions of exponential growth. This,
and a few related approaches, have been provided by Crow and Kimura
(1970, \S 5.3). Specifically, denoting \(\nu_i\) the abundance of
individuals with allele \(i\) and \(m_i\) the growth rate of allele
\(i\) (called the Malthusian parameter in Crow and Kimura 1970), we have
\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation} Starting from this model, we get the total abundance of
the population as \(N=\sum_{i=1}^K\nu_i\), the frequency of allele \(i\)
as \(p_i=\nu_i/N\) and the mean fitness of the population as
\(\bar m=\sum_{i=1}^Kp_im_i\). Hence, we can employ some elementary
calculus to derive the dynamics of abundance \(dN/dt\) and the dynamics
of allele frequencies \(dp_1/dt,\dots,dp_K/dt\) as \begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation} \begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation} Two important observations of these equations include (i)
mean fitness \(\bar m\) determines the abundance dynamics of the entire
population and (ii) allele \(i\) will increase (decrease) in frequency
if \(m_i>\bar m\) (\(<\bar m\)). Equation (\ref{popgen}) is known in the
field of evolutionary game theory as the replicator equation (Hofbauer
and Sigmund 1998; Nowak 2006; Lion 2018; Taylor and Jonker 1978;
Schuster and Sigmund 1983). Instead of being explicitly focused on
alleles, the replicator equation describes the fluctuations of relative
abundances of various \emph{types} in a population in terms of the vital
rates of each type. Using a matrix of transition rates between differing
types, it is straight-forward to extend the replicator equation to
include mutation, which is known as the replicator-mutator equation
(Nowak 2006).

@@

\begin{multline}\label{N_det}
\frac{d}{dt}N(t)=\frac{d}{dt}\int_{-\infty}^{+\infty}\nu(x,t)dx=\int_{-\infty}^{+\infty}\frac{\partial}{\partial t}\nu(x,t)dx \\
=\int_{-\infty}^{+\infty} m(\nu,x)\nu(x,t)dx+\int_{-\infty}^{+\infty}\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)dx \\
=N(t)\int_{-\infty}^{+\infty} m(\nu,x)p(x,t)dx=\bar m(t) N(t).
\end{multline} Using our assumptions on mutation and rate of growth, we
show in SM \S\ref{finite} \(\nu(x,t)\) is twice differentiable with
respect to \(x\) and \(\int_{-\infty}^{+\infty}\nu(x,t)dx<\infty\) for
all \(t\geq0\). This implies that we are justified in swapping the order
of differentiation and integration and the result
\(\int_{-\infty}^{+\infty}\frac{\partial^2}{\partial x^2}\nu(x,t)dx=0\)
can be derived from the fundamental theorem of calculus. Biological
reasoning agrees with this latter result since mutation neither creates
nor destroys individuals, but merely changes their type from their
parental type. Taking the same approach, we derive the dynamics of the
phenotypic distribution \(p(x,t)\) in response to selection and muation
as

>>>>

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{bioarxiv} % Sets Journal name

\usepackage{lineno} % add
  \linenumbers % turns line numbering on
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A White Noise Approach to Evolutionary Ecology},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc header
\usepackage{xr}
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{hyperref}
\usepackage{amsmath}
\externaldocument{SM}
\modulolinenumbers[2]
\usepackage{float}
\usepackage{flafter}

\begin{document}
\begin{frontmatter}

  \title{A White Noise Approach to Evolutionary Ecology}
    \author[Bio]{Bob Week\corref{1}}
   \ead{bobweek@gmail.com} 
    \author[Stats]{Steve Krone}

    \author[Bio]{Scott L. Nuismer}

    \author[Bio]{Luke J. Harmon}

      \address[Bio]{University of Idaho, Department of Biological Sciences, 875 Perimeter
Drive MS 3051, Moscow, Idaho, 83844}
    \address[Stats]{University of Idaho, Department of Statistical Science, 875 Perimeter
Drive MS 1104, Moscow, Idaho, 83844}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  We derive the dynamics of the distribution of a quantitative character
  and the abundance of a biological population from a stochastic partial
  differential equation driven by space-time white noise. In the process
  we develop a useful set of heuristics to operationalize the powerful,
  but abstract theory of white noise and measure-valued stochastic
  processes. This approach allows us to compute the full implications of
  demographic stochasticity on phenotypic distributions and abundances of
  populations. We demonstrate the utility of our approach by deriving a
  quantitative genetic model of diffuse coevolution mediated by
  exploitative competition for a continuum of resources. Other than trait
  and abundance distributions, this model predicts interaction networks
  parameterized by rates of interactions, competition coefficients, and
  selection gradients. We briefly investigate the relationship between
  selection gradients and competition coefficients. This illustrative
  investigation suggests absolute values of linear selection gradients are
  negatively correlated with competition coefficients when the ratio of
  interspecific trait variation to intraspecific trait variation is
  approximately less than \(1.25\) and positively correlated otherwise.
  The inverse relationship holds for quadratic selection gradients except
  with threshold diversity ratio equal to unity instead of \(1.25\).
  Hence, this approach contributes to the development of a synthetic
  theory of evolutionary ecology by formalizing first principle
  derivations of stochastic models that underlie rigorous investigations
  of the relationship between feedbacks of biological processes and the
  patterns of diversity they produce.
  \end{abstract}

 \end{frontmatter}

@@

Our goal in this manuscript is to develop a rigorous, but accessible
approach to synthesize the stochastic dynamics of abundance, mean trait
and heritable variation in biological populations for the study of
theoretical evolutionary ecology. A primary aim of theoretical
evolutionary ecology is the development of mathematical approaches to
describe the evolution of populations and their interactions with both
the biotic and abiotic environments in which they are embedded. Given
this consideration, a natural scope for such an approach centers on
quantifying the abundance dynamics of populations and the evolution of
traits mediating their interactions as functions of relevant abiotic
factors. Although taking into account abundance, phenotype and
environment provides the basis for a partial understanding of the
complex nature of biological communities, a deeper understanding must
account for the effects of dispersal and the phylogeographic history of
interacting lineages (Hickerson et al., 2010; Kraft et al., 2007;
Manceau et al., 2016; McPeek, 2017) along with the genetic basis of
ecologically relevant traits (Conner, 2004; Fussman et al., 2007) and
feedbacks between populations and the biogeochemical cycles they
ultimately depend on (Ågren and Andersson, 2012; Loreau, 2010). It is
therefore ideal that the development of any such mathematical approach
anticipates extensions to account for these important factors shaping
ecological communities, especially as empirical and conceptual work in
these directions continues to grow (Abdala-Roberts and Mooney, 2014;
Crutsinger, 2015; Fitzpatrick et al., 2015, 2017; Harmon et al., 2019;
Kölzsch et al., 2015; Marx et al., 2017; Nuland et al., 2019; Rudman et
al., 2017; Skovmand et al., 2018). Furthermore, the approach would
benefit from a stochastic component to capture the chance nature of
biological reality (Lande et al., 2003; Meester et al., 2018; Mubayi et
al., 2019) and serve as a basis for the construction of statistical
methods that measure evolutionary and ecological processes occuring in
the wild. Such methods will tether theory to reality and allow for
rigorous tests of hypotheses on the structure and behavior of ecological
communities. In this paper we introduce a framework that establishes a
formal connection between the continuous-time dynamics of abundance and
quantitative traits in stochastically evolving populations. We then
demonstrate the utility of our framework through the derivation and
analysis of a model of diffuse coevolution and discuss how it can be
extended to account for the details mentioned above.

Current theoretical approaches to synthesize evolution and ecology have
capitalized on the fact that biological fitness plays a key role in
determining both sets of dynamics. While correlation of fitness and
genotype is the basis of evolution by natural selection, the mean
fitness across all individuals in a population determines the growth,
stasis or decline of abundance. In section \ref{deterministic} we review
the mathematical formalization of this connection, which has been
established in the contexts of population genetics (Crow and Kimura,
1970; Roughgarden, 1979), evolutionary game theory (Hofbauer and
Sigmund, 1998; Lion, 2018; Nowak, 2006), quantitative genetics (Doebeli,
1996; Lande, 1982; Lion, 2018) and a unifying framework for these three
distinct approaches to evolutionary theory (Champagnat et al., 2006)
which includes our approach as a special case. We note here there are
two primary differences between the results presented in this paper and
those of Champagnat, Ferrière and Méléard (2006). First, instead of
developing a unifying approach to evolutionary theory, we focus on
developing a stochastic synthesis of population dynamics and
quantitative genetics including specific expressions for the dynamics of
abundance, mean trait and additive genetic variance. Second, although
Champagnat, Ferrière and Méléard (2006) provide rigorous proofs of
results relevant to our approach, they do not communicate the nessecary
tools we make use of for deriving the dynamics of abundance, mean trait
and additive genetic variance. We aim to communicate these tools to a
wide audience by striking a balance between intuitive appeal and
mathematical rigor.

Although reviewing previously established theoretical results reveals a
beautiful synthesis of evolution and population ecology, it also reveals
a gap in theoretical approaches to incorporate the intrinsically random
nature of populations. Specifically, in theoretical quantitative
genetics the derivation of a population's response to random genetic
drift is derived in discrete time under the assumption of constant
effective population size using arguments based on properties of random
samples (Lande, 1976). Though this approach conveniently mimics the
formalism provided by the Wright-Fisher model of population genetics,
real population sizes fluctuate over time. Furthermore, since these
fluctuations are themselves stochastic, it seems natural to derive
expressions for the evolutionary response to demographic stochasticity
and consider how the results relate to characterizations of random
genetic drift. This can be done in continuous time for population
genetic models without too much technical overhead, assuming a finite
number of alleles (Gomulkiewicz et al., 2017; Lande et al., 2009;
Parsons et al., 2010). However, for populations with a continuum of
types, such as a quantitative trait, finding a formal approach to derive
the evolutionary response to demographic stochasticity has remained a
vexing mathematical challenge. In this paper we close this gap by
combining the calculus of white noise with results on rescaled limits of
branching Brownian motion processes (BBM) and stochastic partial
differential equations (SPDE), which are stochastic analogs of
deterministic partial differential equations (PDE). Our goal has two
components: 1) Establish a novel synthetic approach to theoretical
evolutionary ecology that provides a formal connection between
demographic stochasticity and random genetic drift in the context of
quantitative traits. 2) Communicate some useful properties of space-time
white noise, BBM and SPDE to as wide of audience as possible. With this
goal in mind we will not provide a rigorous treatment of any of these
deep subjects. Instead, we introduce a set of heuristics that only
require the basic concepts of Reimann integration, partial
differentiation and some exposure to Brownian motion and stochastic
ordinary differential equations (SDE). For a concise introduction to SDE
and Brownian motion, we recommend the primer by Evans (2014). Rigorous
treatments of SPDE and rescaled limits of BBM can be found in Walsh
(1986) and Dawson (1993) respectively.

Through accomplishing these goals this work yields five main results.
Firstly, we develop a concise, but accurate approach for understanding
SPDE and provide heuristics to empower mathematical evolutionary
ecologists to derive novel models that relax key assumptions commonly
made in combining models of evolutionary and abundance dynamics. Second,
we formally derive general SDE driven by demographic stochasticity that
express the joint dynamics of abundance, mean trait and additive genetic
variance for populations exhibiting a wide range of potential phenotypic
distributions. Although these dynamics have been considered before while
accounting for demographic stochasticity (Polechová and Barton, 2015),
to our knowledge this is the first time they have been rigorously
derived from first principles. Third, we simplify these SDE by
approximating the abundance density of phenotypes with a Gaussian curve
and integrate the results with a model of imperfect inheritance to find
formula that generalize classical quantitative genetic expressions. This
includes an expression for the response of additive genetic variance to
selection in terms of fitness gradients analogous to the expression for
the response in mean phenotype found by Lande (1976). Fourth, following
classical niche theory we provide a rigorous mechanistic derivation of
fitness for populations experiencing resource competition and abiotic
stabilizing selection using diffusion limits of measure-valued branching
processes. This calculation provides a demonstration for performing such
diffusion limits that can be applied in other contexts. Fifth, we
explore implications of the resulting model of diffuse coevolution by
investigating the relationship between selection gradients and
competition coefficients, discovering the conditions under which these
quantities positively or negatively covary.

@@

At the core of our approach is a stochastic analog of the replicator
equation with mutation in continuous time and phenotypic space (Schuster
and Sigmund, 1983; Taylor and Jonker, 1978). From this stochastic
replicator-mutator equation we derive a system of SDE for the dynamics
of abundance, mean trait and additive genetic variance of a population.
Hence, our approach develops a quantitative genetic theory of
evolutionary ecology. A popular alternative to quantitative genetics is
the theory of adaptive dynamics. As demonstrated by Page and Nowak
(2002) and Champagnat, Ferrière and Méléard (2006), the canonical
equation of adaptive dynamics can be derived from the replicator-mutator
equation. Thus, one could start from the atomic roots of our approach
and pursue a stochastic adaptive dynamic theory instead. We choose the
former in anticipation of an extension of our approach that explicitly
models the genetic details of populations.

@@

Our review begins by considering the dynamics of an asexually
reproducing population in a homogeneous environment. For simplicity, we
first assume individuals are haploid and carry one of \(K\) alleles each
with a different fitness expressed as growth rate before introducing a
model involving a quantitative trait. Under these assumptions, the
derivation of the evolution of allele frequencies due to natural
selection can be derived from expressions of exponential growth. This,
and a few related approaches, have been provided by Crow and Kimura
(1970, \S 5.3). Specifically, denoting \(\nu_i\) the abundance of
individuals with allele \(i\) and \(m_i\) the growth rate of allele
\(i\) (called the Malthusian parameter in Crow and Kimura, 1970), we
have

\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation}

Starting from this model, we get the total abundance of the population
as \(N=\sum_{i=1}^K\nu_i\), the frequency of allele \(i\) as
\(p_i=\nu_i/N\) and the mean fitness of the population as
\(\bar m=\sum_{i=1}^Kp_im_i\). Hence, we can employ some elementary
calculus to derive the dynamics of abundance \(dN/dt\) and the dynamics
of allele frequencies \(dp_1/dt,\dots,dp_K/dt\) as

\begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation} \begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation}

Two important observations of these equations include (i) mean fitness
\(\bar m\) determines the abundance dynamics of the entire population
and (ii) allele \(i\) will increase (decrease) in frequency if
\(m_i>\bar m\) (\(<\bar m\)). Equation (\ref{popgen}) is known in the
field of evolutionary game theory as the replicator equation (Hofbauer
and Sigmund, 1998; Lion, 2018; Nowak, 2006; Schuster and Sigmund, 1983;
Taylor and Jonker, 1978). Instead of being explicitly focused on
alleles, the replicator equation describes the fluctuations of relative
abundances of various \emph{types} in a population in terms of the vital
rates of each type. Using a matrix of transition rates between differing
types, it is straight-forward to extend the replicator equation to
include mutation, which is known as the replicator-mutator equation
(Nowak, 2006).

@@

\begin{multline}\label{N_det}
\frac{d}{dt}N(t)=\frac{d}{dt}\int_{-\infty}^{+\infty}\nu(x,t)dx=\int_{-\infty}^{+\infty}\frac{\partial}{\partial t}\nu(x,t)dx \\
=\int_{-\infty}^{+\infty} m(\nu,x)\nu(x,t)dx+\int_{-\infty}^{+\infty}\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)dx \\
=N(t)\int_{-\infty}^{+\infty} m(\nu,x)p(x,t)dx=\bar m(t) N(t).
\end{multline}

Using our assumptions on mutation and rate of growth, we show in SM
\S\ref{finite} \(\nu(x,t)\) is twice differentiable with respect to
\(x\) and \(\int_{-\infty}^{+\infty}\nu(x,t)dx<\infty\) for all
\(t\geq0\). This implies that we are justified in swapping the order of
differentiation and integration and the result
\(\int_{-\infty}^{+\infty}\frac{\partial^2}{\partial x^2}\nu(x,t)dx=0\)
can be derived from the fundamental theorem of calculus. Biological
reasoning agrees with this latter result since mutation neither creates
nor destroys individuals, but merely changes their type from their
parental type. Taking the same approach, we derive the dynamics of the
phenotypic distribution \(p(x,t)\) in response to selection and muation
as

\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=\Big(m(\nu,x)-\bar m(t)\Big) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}

This result closely resembles Kimura's continuum-of-alleles model
(Bürger, 1986; Kimura, 1965). The primary difference being that our
model utilizes diffusion instead of convolution with an arbitrary
mutation kernel. Of course, our model of mutation can be derived as an
approximation to Kimura's model, which, for sexually reproducing
populations, has been referred to as the Gaussian allelic approximation
in reference to the distribution of mutational effects at loci in a
genome on the values of traits (Lande, 1975; but see also Bürger, 2000;
along with Johnson and Barton, 2005), the infinitesimal model in
reference to modelling continuous traits as being encoded by an infinite
number of loci each having infinitesimal effect (Barton et al., 2017)
and the Gaussian descendants approximation in reference to offspring
trait values being normally distributed around their parental values
(Turelli, 2017). Alternatively, since diffusion is the continuous-time
equivalent to convolution against a Gaussian kernel (SM
\S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a
special case of Kimura's continuum-of-alleles model. We return to these
concepts in \S\ref{inheritance} where we review a simple model of
inheritance that leads us to an expression for the stochastic evolution
of additive genetic variance. To distinguish this model from previous
models of phenotypic evolution we refer to PDE (\ref{eq1}) from which
(\ref{p_PDE}) was derived as the Deterministic Asexual Gaussian allelic
model with Abundance dynamics (abbreviated DAGA). Later, we will extend
this model to include the effects of demographic stochasticity, which we
refer to as the Stochastic Asexual Gaussian allelic model with Abundance
dynamics (abbreviated SAGA).

The covariance of fitness and phenotype across the population is defined
as

\begin{equation}
\mathrm{Cov}_t\Big(m(\nu,x),x\Big)=\int_{-\infty}^{+\infty}\Big(m(\nu,x)-\bar m(t)\Big)\Big(x-\bar x(t)\Big)p(x,t)dx.
\end{equation}

Hence, the dynamics of the mean trait \(\bar x(t)\) can be derived as
